---
title: "Gapminder"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an
independent organization that seeks to educate people about the state of the
world. They seek to counteract the worldview constructed by a hype-driven media
cycle, and promote a "fact-based worldview" by focusing on data. The dataset
we'll study in this challenge is from Gapminder.

# Guided EDA
<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### __q0__ Perform your "first checks" on the dataset. What variables are in this
dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
glimpse(gapminder)
```

**Observations**:

- Write all variable names here
country, continent, year, life expectancy (lifeExp), population (pop), and GDP per capita (gdpPercap)

### __q1__ Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- gapminder %>%
  pull(year) %>%
  max()
year_min <- gapminder %>%
  pull(year) %>%
  min()
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### __q2__ Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(year == year_min) %>%
  ggplot(aes(x = continent, y = gdpPercap))+
  geom_boxplot() + 
  scale_y_log10()
```

**Observations**:

- The dataset states that the median GDPs for each continent go in the following order: Africa, Asia, the Americas, Europe, and Oceania.
- Because we used a boxplot, we can see (roughly) the amount of variation in these datasets: Asia seems to have GDPs that span the widest range, including an outlier that is an order of magnitude ahead of most other continents 
- Oceania has barely any variation at all in its data, indicating either very homogenous or very little data.  

**Difficulties & Approaches**:

- The outlier in Asia is quite dramatic, making most visualization pretty ineffective without any additional axis scaling. However, I searched up code that logarithmically scales the y axis and so that issue is remedied.

### __q3__ You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
outliers_americas <- gapminder %>%
  filter(
    continent == "Americas",
    gdpPercap > 2e+04
    ) 
outliers_asia <- gapminder %>%
  filter(
    continent == "Asia",
    gdpPercap > 5e+04
  )

outliers_americas
outliers_asia
```

**Observations**:

- Identify the outlier countries from q2
  - USA, Canada, and Kuwait.

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### __q4__ Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(year == year_min) %>%
  ggplot(aes(x = continent, y = gdpPercap))+
  geom_boxplot() + 
  geom_point(
    data = . %>% filter(country %in% c("Kuwait", "United States", "Canada")),
    mapping = aes(color = country),
    size = 2
  ) + 
  scale_y_log10()
```

**Observations**:

- Write your observations here

# Your Own EDA
<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### __q5__ Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
gapminder %>%
  filter(year == year_min) %>%
  ggplot(aes(x = continent, y = lifeExp)) +
  geom_boxplot() + 
  geom_point(
    data = . %>% filter(country %in% c("Turkey")),
    mapping = aes(color = country),
    size = 2
  )
  
```

- Plotting life expectancy for each continent yields an interesting outlier in Turkey. Turkey is so far off from the rest of the box plot that it's hard to believe that it could happen naturally. Unsurprisingly, global factors played into this - Turkey acted as a middle point between the Soviets and the United States, and inter-communal violence occurred (and is still occuring) which leads to an understandably lower life expectancy.

```{r q5-task2}
## TASK: Your second graph
gapminder %>%
  ggplot(aes(x = gdpPercap, y = lifeExp, color = year))+
  geom_point() 


```

- If we look at this plot of GDP per capita and life expectancy over time, we see a general trend that later years have higher life expectancy and higher GDP. We can attribute higher life expectancy to things like advancement in medicine, hygiene, etc. And higher GDP to a global economy that's becoming more and more interconnected and thus robust. 

```{r q5-task3}
## TASK: Your third graph
gapminder %>%
  filter(country == "China") %>%
  ggplot(aes(x = year, y = lifeExp, color = country)) +
  geom_line()
# filtered <- gapminder %>%
#   filter(year == 1992) %>%
#   arrange(lifeExp)
# 
# filtered
```

- We see when plotting life expectancy versus time in China that there's a general upward trend except for the dip that occurs around 1960. Further analysis of history suggests that this was likely due to "The Great Leap Forward," an initiative in 1958 that sparked a nationwide famine that resulted in the deaths of many millions of people. Unsurprisingly, life expectancy decreased in that time. 
